---
title: "Kenya External Public Debt Register"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.width = 9,
    fig.height = 5.5,
    dpi = 120
)

library(data.table)
library(ggplot2)
library(scales)
library(knitr)

source("R/analysis_functions.R")

debt <- fread("data/processed/register_2025_ksh_loans.csv")
debt[, agreement_date := as.Date(agreement_date)]
debt[, maturity_date := as.Date(maturity_date)]

reporting_date <- as.Date("2025-06-30")
debt <- add_presidential_period(debt)
debt <- add_maturity_bucket(debt, reporting_date)
debt[, duration_years := as.numeric(maturity_date - agreement_date) / 365.25]

active_debt <- debt[outstanding_june_2025_ksh > 0]
total_outstanding <- sum(debt$outstanding_june_2025_ksh, na.rm = TRUE)
total_repaid <- sum(debt$principal_repaid_ksh, na.rm = TRUE)
total_drawdowns <- sum(debt$drawdowns_ksh, na.rm = TRUE)
active_records <- nrow(active_debt)
active_loans <- uniqueN(active_debt$loan_id)
```

This project presents Kenya's loan-level external public debt recorded by the
National Treasury at **30 June 2025**. All debt-stock and fiscal-year movement
charts are reported in **Kenya shillings (KES)** using the Treasury's published
KSh values; the analysis does not apply a separate exchange-rate conversion.

Source: [National Treasury External Public Debt Register](https://www.treasury.go.ke/external-public-debt-register).

## Headline position

```{r headline-table}
headline <- data.table(
    Indicator = c(
        "External debt outstanding",
        "Active loan records",
        "Distinct active loan IDs",
        "Principal repaid during FY2024/25",
        "Drawdowns during FY2024/25"
    ),
    Value = c(
        kes_trillions(total_outstanding),
        comma(active_records),
        comma(active_loans),
        kes_billions(total_repaid),
        kes_billions(total_drawdowns)
    )
)

kable(headline, align = c("l", "r"))
```

The loan-level closing balances total **`r kes_trillions(total_outstanding)`**.
They differ from the printed Treasury total by only KSh 11, within the documented
KSh 100 rounding tolerance used by this project.

## Outstanding debt by creditor category

```{r debt-by-creditor-category}
category_stock <- active_debt[
    , .(outstanding_ksh = sum(outstanding_june_2025_ksh, na.rm = TRUE)),
    by = creditor_category
][order(outstanding_ksh)]

plot_kes_bar(
    category_stock,
    x = reorder(creditor_category, outstanding_ksh),
    y = outstanding_ksh,
    title = "Multilateral lenders hold the largest share of external debt",
    subtitle = "Outstanding principal at 30 June 2025"
)
```

```{r category-share-table}
category_table <- category_stock[
    order(-outstanding_ksh),
    .(
        `Creditor category` = creditor_category,
        `Outstanding (KES bn)` = round(outstanding_ksh / 1e9, 1),
        `Share of total` = percent(
            outstanding_ksh / total_outstanding,
            accuracy = 0.1
        )
    )
]
kable(category_table, align = c("l", "r", "r"))
```

## Largest creditors

```{r largest-creditors}
top_creditors <- active_debt[
    , .(outstanding_ksh = sum(outstanding_june_2025_ksh, na.rm = TRUE)),
    by = creditor_name
][order(-outstanding_ksh)][1:12]

plot_kes_bar(
    top_creditors[order(outstanding_ksh)],
    x = reorder(creditor_name, outstanding_ksh),
    y = outstanding_ksh,
    title = "The twelve largest external creditors",
    subtitle = "Outstanding principal at 30 June 2025",
    fill = "#1B9E77"
)
```

## Currency exposure measured in KES

The currency identifies the denomination of each loan. The bars show the KES
value of the outstanding stock, not a sum of incompatible foreign-currency
amounts.

```{r currency-exposure}
currency_stock <- active_debt[
    , .(outstanding_ksh = sum(outstanding_june_2025_ksh, na.rm = TRUE)),
    by = currency
][order(outstanding_ksh)]

plot_kes_bar(
    currency_stock,
    x = reorder(currency, outstanding_ksh),
    y = outstanding_ksh,
    title = "Currency composition of outstanding external debt",
    subtitle = "KES value at 30 June 2025",
    fill = "#7570B3"
)
```

## Repayments and drawdowns during FY2024/25

```{r annual-flows}
annual_flows <- data.table(
    movement = factor(
        c("Principal repaid", "Drawdowns"),
        levels = c("Principal repaid", "Drawdowns")
    ),
    amount_ksh = c(total_repaid, total_drawdowns)
)

ggplot(annual_flows, aes(movement, amount_ksh, fill = movement)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    scale_fill_manual(values = c("Principal repaid" = "#D95F02", "Drawdowns" = "#386CB0")) +
    scale_y_continuous(labels = label_number(scale = 1e-9, suffix = " bn"),
                       expand = expansion(mult = c(0, 0.08))) +
    labs(
        title = "Principal repayments exceeded drawdowns during FY2024/25",
        x = NULL, y = "Amount (KES billions)"
    ) +
    external_debt_theme()
```

These are fiscal-year movements reported by Treasury. They should not be read as
the complete change in the KES debt stock because valuation and prior-period
adjustments can also change the closing balance.

## Maturity profile

```{r maturity-profile}
maturity_stock <- active_debt[
    , .(outstanding_ksh = sum(outstanding_june_2025_ksh, na.rm = TRUE)),
    by = maturity_bucket
][order(maturity_bucket)]

maturity_stock[, maturity_bucket := factor(
    maturity_bucket,
    levels = c("Matured", "Within 1 year", "1-3 years", "3-5 years",
               "5-10 years", "More than 10 years")
)]

ggplot(maturity_stock, aes(maturity_bucket, outstanding_ksh)) +
    geom_col(fill = "#E6AB02", width = 0.7) +
    scale_y_continuous(labels = label_number(scale = 1e-9, suffix = " bn"),
                       expand = expansion(mult = c(0, 0.08))) +
    labs(
        title = "Maturity profile of debt outstanding at June 2025",
        subtitle = "Based on contractual maturity dates in the register",
        x = NULL, y = "Outstanding debt (KES billions)"
    ) +
    external_debt_theme() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Contractual loan duration by presidential period

The histogram measures the time between each loan's agreement date and contractual
maturity date. It does not show the years remaining after June 2025. Presidential
periods classify records by agreement date using non-overlapping terms.

```{r duration-by-president, fig.height=7}
plot_duration_histogram(
    debt[!is.na(president)],
    facet = president,
    title = "Contractual loan duration by presidential agreement period",
    subtitle = "Five-year bins; each observation is one loan record",
    columns = 2
)
```

## Contractual loan duration by creditor category

```{r duration-by-creditor-category, fig.height=7}
plot_duration_histogram(
    debt,
    facet = creditor_category,
    title = "Contractual loan duration by creditor category",
    subtitle = "Five-year bins; vertical scales vary by category",
    columns = 3
)
```
## Outstanding stock associated with presidential periods

This classification uses the agreement date and non-overlapping presidential
terms. It shows **June 2025 outstanding stock associated with agreements signed
in each period**. It does not claim that a president personally borrowed the
amount, nor does it measure total disbursements received during a presidency.

```{r presidential-periods}
president_stock <- active_debt[
    !is.na(president),
    .(outstanding_ksh = sum(outstanding_june_2025_ksh, na.rm = TRUE)),
    by = president
][order(outstanding_ksh)]

plot_kes_bar(
    president_stock,
    x = reorder(president, outstanding_ksh),
    y = outstanding_ksh,
    title = "Current stock by period in which agreements were signed",
    subtitle = "Outstanding at 30 June 2025; classification based on agreement date",
    fill = "#386CB0"
)
```

## Methodology and validation

The pipeline extracts only pages **3-29**, whose repeated heading is
`EXTERNAL PUBLIC DEBT REGISTER AS AT END JUNE 2025 (KSHS)`. Pages 1-2 contain
summaries and pages 30-61 contain the separate FX register.

PDF words are assigned to columns using their page coordinates. This avoids
mixing wrapped loan titles, creditor names and two-line categories. The output
retains the source page and row for every record.

The validation checks found:

- 1,041 records and 1,041 distinct loan IDs;
- no duplicate loan/currency keys;
- no missing required fields or invalid dates/currencies;
- no unexpected creditor or borrower categories; and
- a KSh 11 difference from the printed closing total.

See [DATA_PIPELINE.md](DATA_PIPELINE.md) for extraction instructions and detailed
validation evidence.

## Limitations

- The register reports principal-related stock and movements; it is not a full
  fiscal-cost dataset and does not provide loan-level interest payments here.
- A row is treated as a loan record and IDs are counted distinctly. Some economic
  facilities may still contain multiple tranches or currency components.
- Agreement-date classifications do not capture when a loan was negotiated,
  approved, disbursed or ultimately spent.
- KES values can change because of exchange-rate valuation as well as borrowing
  and repayment flows.
